EBASE manuscript

### 2023-03-27 manu-draft.qmd
updating figures with nse results

where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each metabolic parameter, each unique combination of priors, and each optimization period. The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for one day, $n \approx 52$ for seven days, and $n \approx 12$ for 30 days. All metabolic estimates were averaged within the number of days for each optimization period prior to calculating $NSE$ values, if appropriate. The $NSE$ value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1.  Values in the positive range from 0 to 1 are desirable with 1 being a perfect fit, whereas values in the negative range indicate the mean of the synthetic data is a better predictor than EBASE.

@@

Several conclusions can be made from the results in @fig-priorcomp based on the patterns of the $R^2$ values. Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters. Changing the variance of the $a$ parameter had little effect on any of the comparisons. Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, although the precision was worse for the seven day optimization period (e.g., the $a$ parameter). Conversely, larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the seven day optimization period.

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combination of prior distributions, where the means and standard deviations are evaluated from low (L) to high (H) for a one day and seven day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and synthetic values. The $b$ parameter is a fixed constant and was not evaluated.  Rankings of each combination as the average $R^2$ value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

@@

![Add caption.](figs/priorsumcomp.png){#fig-priorsumcomp}

@@

![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters. The best and worst performing models for the one day (a, b) and seven day (c, d) optimization periods are based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-priorcomp).  The results are averaged at the time step of the model optimization period. The $b$ parameter is fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.](figs/optex.png){#fig-optex fig-align="center"}

>>>>

where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each metabolic parameter, each unique combination of priors, and each optimization period. The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for one day, $n \approx 52$ for seven days, and $n \approx 12$ for 30 days. All metabolic estimates were averaged within the number of days for each optimization period prior to calculating NSE values, if appropriate. The NSE value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1.  Values in the positive range from 0 to 1 are desirable with 1 being a perfect fit, whereas values in the negative range indicate the mean of the synthetic data is a better predictor than EBASE.

@@

Several conclusions can be made from the results in @fig-priorcomp based on the patterns of the NSE values. Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters. Changing the variance of the $a$ parameter had little effect on any of the comparisons. Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, although the precision was worse for the seven day optimization period (e.g., the $a$ parameter). Conversely, larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the seven day optimization period.

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combination of prior distributions, where the means and standard deviations are evaluated from low (L) to high (H) for a one day, seven day, and thirty day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate the mean of the synthetic data is a better predictor than EBASE.  The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display. The $b$ parameter is a fixed constant and was not evaluated.  Rankings of each combination as the median NSE value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

@@

![Summaries of median Nash-Sutcliffe Efficiency values across all parameters for the comparisons of EBASE and synthetic model outputs separated by different combinations of prior values. Each result on the x-axis represents the median values for which the combination prior applies, e.g., all results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.  Point size represents the inter-quartile range of median NSE values for each result.](figs/priorsumcomp.png){#fig-priorsumcomp}

@@

![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters. The best and worst performing models for the one day (a, b), seven day (c, d), and thirty day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficenciy values across all parameter estimates (@fig-priorcomp).  The results are averaged at the time step of the model optimization period. The $b$ parameter is fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.](figs/optex.png){#fig-optex fig-align="center"}
